---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# GCFPCA

<!-- badges: start -->
[![pkgdown website](https://img.shields.io/badge/docs-pkgdown-blue)](https://yulu98.github.io/GCFPCA/) 
<!-- badges: end -->

Generalized Conditional Functional Principal Components Analysis (GCFPCA) is an R package for performing functional principal component analysis under generalized exponential family distributions.

* **Authors**: [Yu Lu](mailto:ylu136@jhu.edu), Xinkai Zhou, Erjia Cui, Dustin Rogers, Ciprian Crainiceanu, Julia Wrobel, Andrew Leroux
* **Version**: 0.9

## Installation

You can install the development version of GCFPCA from GitHub:

```{r, echo = TRUE, eval = FALSE}
# install.packages("devtools")
devtools::install_github("yulu98/GCFPCA", build_vignettes = TRUE)
```

## Documentation and Examples

For detailed examples and usage of the GCFPCA package, please refer to the vignette available on the [pkgdown website](https://yulu98.github.io/GCFPCA/articles/gcfpca.html).
