---
title: "GC-FPCA: a vignette"
author: "Yu Lu <ylu136@jh.edu>"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GC-FPCA: a vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  cache = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.width = 6
)
```



The `GCFPCA` package is for the joint modeling of the fixed and random effects of non-Gaussian functional outcomes. This vignette outlines the general functionality of the package. The main function in the package is called `gcfpca()`. An additional function called `gcfpca_simu()` can be used to simulate binary, count, or normally distributed functional observations. 

Currently, the package is designed to handle only dense functional data where each subject is observed on the same grid of regularly spaced points along the domain. In the future the package will be expanded to handle sparse and irregular functional observations.


# The `GC-FPCA` model and algorithm 

The model for generalized functional principal components analysis is 

$$
g\left(E[Z_i(s) | X_1, \cdots, X_p]\right) = \beta_0(s) + \sum_{r=1}^{p} X_{ir} \beta_r(s) + \sum_{l=1}^{L} \xi_{il} \phi_l(s)
$$


where $Z_i(s)$ is the observed data at time $s$ for subject $i$, $\beta_0(s)$ and $\beta_r(s)$ are domain-varying coefficient, $\xi_{ik}\sim N(0, \lambda_k)$ are mutually independent subject-specific scores, and $\phi_k(s); k\in 1,\ldots,K$ are the eigenfunctions.

# The `GCFPCA` package

## A note on data formatting

Like most model-fitting functions in R, lmer takes as its first two arguments a *formula* specifying the model and the *data* with which to evaluate the formula. The *formula* use the same format as required in `lme4:lmer`. Use of this package requires that *data* be in a specific format: a long-form data frame with variables `id`, `index`, `Y`, and fixed effect predictors, where the `Y` column contains functional observations for all subjects, the `id` column identifies which observations belong to which subject, and `index` provides the grid (domain) over which the `Y`s are observed, and fixed effect predictors are not domain-dependent, but only repeated for each time point.

The variable `id` should be a unique identifier in that each id identifies a single subject. Since we assume there is only one curve per subject for this package, `id` uniquely identifies each curve as well. Other predictors can be included in the data as long as the variables `id`, `index`, and `Y` are present.

```{r load_libraries, message = FALSE}
library(GCFPCA)
library(tidyverse)
library(lme4)
library(refund)
library(mvtnorm)
library(splines)
theme_set(theme_minimal())
```

## Simulate data for GC-FPCA

`gcfpca_simu()` generates curves with a population mean, a fixed effect (binary or gaussian distributed) $K = 4$ functional principal components, and true eigenvalues $\lambda_k = 0.5^{k-1}; k = 1,2,3,4$. When function argument `fe_case = 1`, the fixed effect predictor is $Bernoulli(\frac12)$ distributed. When `fe_case = 2`, the fixed effect predictor is $N(0, 1)$ distributed. When function argument `re_case = 1`, the true eigenfunctions are periodic and defined by $\boldsymbol{\phi}(s) = \{\sqrt{2}sin(2\pi s), \sqrt{2}cos(2\pi s), \sqrt{2}sin(4\pi s), \sqrt{2}cos(4\pi s)\}$. When `re_case = 2`, eigenfunctions are non-periodic and given by $\boldsymbol{\phi}(s) = \{1, \sqrt{3}(2s-1), \sqrt{5}(6s^2-6s + 1), \sqrt{7}(20s^3-30s^2+12s-1)\}.$


The argument `family` defines the outcome distribution of the observed data.The code below generates binary functional data with $I = 100$ subjects and $K = 200$ grid points with binary fixed effect predictor and periodic eigenfunctions.

```{r sim_data}
# simulate data

I = 100
K = 200

# simulate data and save results
set.seed(1134)

#set true fixed effects
theta_0 = rnorm(10 + 4, sd=1)
theta_1 = rnorm(10 + 4, sd=1)

#set true fixed effects
f_0 = function(s) splines::bs(s, knots=seq(0.1, 0.9, len=10), Boundary.knots=c(0,1), intercept=TRUE) %*% theta_0
f_1 = function(s) splines::bs(s, knots=seq(0.1, 0.9, len=10), Boundary.knots=c(0,1), intercept=TRUE) %*% theta_1

bin_sim <- gcfpca_simu(I = I, K = K, family = "binomial",
                         beta0_true = f_0, beta1_true = f_1)

ls(bin_sim)
```

The resulting data object,`simu_data`, is a list with values `df_gcfpca`, `lambda`, `phi`, and `scores`.  `simu_data$df_gcfpca` is a data frame with variables `id`, `index`, `Y`, and `X`, which is consistent with the format our `GCFPCA` software requires. `id` is the identifier for a particular subject, `index` represents the functional domain point, the `Y` variable contains binary observations, `X` is the subject-specific fixed effect predictor. 


## Binary outcomes

`gcfpca()` is the main function for the `GCFPCA` package. The key arguments for this function are listed below.

* `formula` A formula specifying the GLMM model
* `data` A data frame with columns: `id` (subject), `index` (time or index), and `Y` (outcome variable).
* `binwidth` determines the width of bins. Should be a positive integer value of $\le J/10$.
* `npc` indicates the number of functional principal components to use. The latter can also be chosen based on an explained share of variance, see argument `pve`.
* `family` is used to specify what type of exponential family functional distribution and link function you would like to use for your data. This parameter is fed to the the functions `glm::glmer()` and `mgcv::bam()`; see the help files for these functions for which types of families and link functions are supported.

The following code uses a binomial family with `logit` link.

```{r gcfpca_bin, message = FALSE, warning=FALSE}

# run fast gfpca model
bin_model <- gc_fpca(formula = Y ~ X + (1|id),
                      data = bin_sim$df_gcfpca,
                      binwidth = 10,
                      family = "binomial",
                      pve = 0.95, npc = 4)
```

The plot below show the estimated fixed effect coefficients for the `logit` binary FPCA models.

```{r plot_bFPCA_fe, echo = TRUE, fig.show='hold'}
plot_df <- cbind.data.frame(sind = rep(seq(0, 1, length.out = K), 2),
                            betahat = c(data.matrix(bin_model$betaHat)),
                            betatrue = c(f_0(seq(0, 1, length.out = K)),
                                         f_1(seq(0, 1, length.out = K))),
                            X = c(rep("Intercept", K),
                                  rep("X", K)),
                            CI_L_pw = c(data.matrix(bin_model$CI_L_pw)),
                            CI_U_pw = c(data.matrix(bin_model$CI_U_pw)),
                            CI_L_joint = c(data.matrix(bin_model$CI_L_joint)),
                            CI_U_joint = c(data.matrix(bin_model$CI_U_joint))) %>%
  mutate(X = factor(X, levels = c("Intercept",
                                  "X")))

plot_df %>%
  ggplot(aes(x = sind, y = betahat)) +
  geom_ribbon(aes(ymin = CI_L_joint, ymax = CI_U_joint, fill = "CI Joint"), alpha = 0.5) +
  geom_ribbon(aes(ymin = CI_L_pw, ymax = CI_U_pw, fill = "CI"), alpha = 0.5) +
  geom_line(aes(color = "GCFPCA")) +
  geom_line(aes(x = sind, y = betatrue, color = "truth")) +
  scale_fill_manual(values = c("CI" = "black", "CI Joint" = "lightgray"), name = "Confidence Interval") +
  scale_color_manual(values = c("GCFPCA" = "darkblue", "truth" = "red"), name = "Confidence Interval") +
  # Adding a horizontal dotted line at y = 0
  geom_hline(yintercept = 0, linetype = "dotted") +
  # Setting x-axis labels to show time
  scale_x_continuous(breaks = seq(0, 1, by = 0.2)) +
  # Facet the plot by variable X, with 3 columns
  facet_wrap(~X, ncol = 3, scales = "free_y") +
  # Adding axis labels and title
  labs(x = "Functional Domain", y = "", fill = "Confidence Interval")
```
The plot below show the estimated principal components for the `logit` binary FPCA models. 

```{r plot_bFPCA_re, echo = TRUE, fig.show='hold'}

flip_efunctions = function(data){
  if(data[["logit_mod"]] %*% data[["truth"]] < 0){
    data[["logit_mod"]] = -1 * data[["logit_mod"]]
  }
  data
}

phi_df = tibble(s = rep(seq(0, 1, length.out  = 200), 4),
                l = rep(paste0("eigenfunction ",1:4), each = 200),
                truth = c(bin_sim$phi),
                logit_mod = c(data.matrix(bin_model$efunctions[, -1]))) %>%
  nest_by(l) 

new_phi = map(phi_df$data, flip_efunctions)
phi_df$new_phi <- new_phi

phi_df %>%
  select(-data) %>%
  unnest(new_phi) %>%
  ungroup() %>% 
  mutate(logit_mod = logit_mod * sqrt(200)) %>%
  pivot_longer(truth:logit_mod, names_to = "model", values_to = "value") %>%
  ggplot(aes(s, value, group = model, color = model, linetype = model)) +
  geom_line() +
  facet_wrap(~l)
```

The plot below shows latent subject-specific means on the linear predictor scale, $\beta_0(s) + \sum_{r=1}^p X_{ir} \beta_r(s) + \sum_{l=1}^L \xi_{il} \phi_l(s)$, for three subjects in the simulated Poisson dataset. True values are black solid lines while GCFPCA estimates are pink dotted lines.


```{r plot_bFPCA_eta}
bin_sim$df_gcfpca %>%
  mutate(eta_hat = as.vector(bin_model$etas)) %>%
  filter(id %in% c(1, 2, 8)) %>%
  ggplot(aes(index, eta)) +
  geom_line() +
  geom_line(aes(y = eta_hat), linetype = 2, color = "salmon") +
  facet_wrap(~id)
```

## Poisson outcomes

The `gcfpca()` function can also perform Poisson GCFPCA on count outcomes. The code below simulates Poisson functional data with gaussian distributed fixed effect predictor and non-periodic eigenfunctions, and performs Poisson GCFPCA. 

```{r pois_mod, message=FALSE, warning=FALSE}
# simulate data
library(splines)

I = 500
K = 1000

# simulate data and save results
set.seed(1134)

#set true fixed effects
theta_0 = rnorm(10 + 4, sd=1)
theta_1 = rnorm(10 + 4, sd=1)

#set true fixed effects
f_0 = function(s) bs(s, knots=seq(0.1, 0.9, len=10), Boundary.knots=c(0,1), intercept=TRUE) %*% theta_0
f_1 = function(s) bs(s, knots=seq(0.1, 0.9, len=10), Boundary.knots=c(0,1), intercept=TRUE) %*% theta_1

count_sim <- gcfpca_simu(I = I, K = K, family = "poisson",
                         beta0_true = f_0, beta1_true = f_1,
                         fe_case = 2, re_case = 2)

count_model <- gc_fpca(formula = Y ~ X + (1|id),
                      data = count_sim$df_gcfpca,
                      binwidth = 20,
                      family = "poisson",
                      pve = 0.95, npc = 4, periodicity = FALSE)
```

The plot below show the estimated fixed effect coefficients for the poisson FPCA models.

```{r plot_pFPCA_fe, echo = TRUE, fig.show='hold'}
plot_df <- cbind.data.frame(sind = rep(seq(0, 1, length.out = K), 2),
                            betahat = c(data.matrix(count_model$betaHat)),
                            betatrue = c(f_0(seq(0, 1, length.out = K)),
                                         f_1(seq(0, 1, length.out = K))),
                            X = c(rep("Intercept", K),
                                  rep("X", K)),
                            CI_L_pw = c(data.matrix(count_model$CI_L_pw)),
                            CI_U_pw = c(data.matrix(count_model$CI_U_pw)),
                            CI_L_joint = c(data.matrix(count_model$CI_L_joint)),
                            CI_U_joint = c(data.matrix(count_model$CI_U_joint))) %>%
  mutate(X = factor(X, levels = c("Intercept",
                                  "X")))

plot_df %>%
  ggplot(aes(x = sind, y = betahat)) +
  geom_ribbon(aes(ymin = CI_L_joint, ymax = CI_U_joint, fill = "CI Joint"), alpha = 0.5) +
  geom_ribbon(aes(ymin = CI_L_pw, ymax = CI_U_pw, fill = "CI"), alpha = 0.5) +
  geom_line(aes(color = "GCFPCA")) +
  geom_line(aes(x = sind, y = betatrue, color = "truth")) +
  scale_fill_manual(values = c("CI" = "black", "CI Joint" = "lightgray"), name = "Confidence Interval") +
  scale_color_manual(values = c("GCFPCA" = "darkblue", "truth" = "red"), name = "Confidence Interval") +
  # Adding a horizontal dotted line at y = 0
  geom_hline(yintercept = 0, linetype = "dotted") +
  # Setting x-axis labels to show time
  scale_x_continuous(breaks = seq(0, 1, by = 0.2)) +
  # Facet the plot by variable X, with 3 columns
  facet_wrap(~X, ncol = 3, scales = "free_y") +
  # Adding axis labels and title
  labs(x = "Functional Domain", y = "", fill = "Confidence Interval")
```
The plot below show the estimated principal components for the poisson FPCA models. 

```{r plot_pFPCA_re, echo = TRUE, fig.show='hold'}
flip_efunctions = function(data){
  if(data[["logit_mod"]] %*% data[["truth"]] < 0){
    data[["logit_mod"]] = -1 * data[["logit_mod"]]
  }
  data
}

phi_df = tibble(s = rep(seq(0, 1, length.out  = K), 4),
                l = rep(paste0("eigenfunction ",1:4), each = K),
                truth = c(count_sim$phi),
                logit_mod = c(data.matrix(count_model$efunctions[, -1]))) %>%
  nest_by(l) 

new_phi = map(phi_df$data, flip_efunctions)
phi_df$new_phi <- new_phi

phi_df %>%
  select(-data) %>%
  unnest(new_phi) %>%
  ungroup() %>% 
  mutate(logit_mod = logit_mod * sqrt(K)) %>%
  pivot_longer(truth:logit_mod, names_to = "model", values_to = "value") %>%
  ggplot(aes(s, value, group = model, color = model, linetype = model)) +
  geom_line() +
  facet_wrap(~l)
```

The plot below shows latent subject-specific means on the linear predictor scale, $\beta_0(s) + \sum_{r=1}^p X_{ir} \beta_r(s) + \sum_{l=1}^L \xi_{il} \phi_l(s)$, for three subjects in the simulated Poisson dataset. True values are black solid lines while GCFPCA estimates are pink dotted lines.


```{r plot_pFPCA_eta}
count_sim$df_gcfpca %>%
  mutate(eta_hat = as.vector(count_model$etas)) %>%
  filter(id %in% c(1, 2, 8)) %>%
  ggplot(aes(index, eta)) +
  geom_line() +
  geom_line(aes(y = eta_hat), linetype = 2, color = "salmon") +
  facet_wrap(~id)
```


## Real data

Here we subsetting the NHANES data to show an example of how to apply our method on real life data. The processed single level NHANES physical activity data can be found on the website of Functional Data Analysis with R book website (https://functionaldataanalysis.org/dataset_nhanes.html). This dataset is also included here in this packge as `nhanes_fda_with_r.rds`. 

We deleted all the missingness and subsetted 50 subjects to show our method.

```{r nhanes_process, echo=FALSE, eval=FALSE}
df_single <- readRDS("/Users/luyu/Library/CloudStorage/OneDrive-JohnsHopkins/WITgroup/GCFPCA/data/nhanes_fda_with_r.rds") %>%
  na.omit()

set.seed(123)  # Set seed for reproducibility
selected_ids <- sample(df_single$SEQN, 50)

# Filter the dataset for the selected IDs
df_selected <- df_single %>% 
  filter(SEQN %in% selected_ids) %>%
  mutate(mims = ifelse(MIMS > 10.558, 1, 0))
dmims <- df_selected$mims[, 1:1440]


# Convert the 'MIMS' matrix to long format
df_long <- df_selected %>%
  select(SEQN, age, gender) %>%
  bind_cols(dmims) %>%
  rename(id = SEQN) %>%
  pivot_longer(cols = MIN0001:MIN1440,
               names_to = "MIN",
               values_to = "mims") %>%
  mutate(MIN = gsub("^MIN*", "", MIN)) %>%
  mutate(MIN = as.numeric(MIN)) %>%
  mutate(gender = ifelse(gender == "Male", 0, 1)) %>%
  rename(index = MIN) %>%
  mutate(id = factor(id))
saveRDS(df_long, "../data/df_long.rds")
```

```{r df_long}
df_long <- readRDS("../data/df_long.rds")
```

First, we fit our GCFPCA model on the subsetted data. ${\rm logit}\{P\{Z_i(s)=1\}\} = \beta_0(s) + \beta_1(s) \text{Age}_i + \beta_2(s) \text{Gender}_i + \sum_{l=1}^{4} \xi_{il}\phi_l(s)$

```{r nhanes_gcfpca, warning=FALSE, message=FALSE}
gcfpca_start_t = Sys.time()
nhanes_gcfpca <- gc_fpca(formula = mims ~ age + gender +(1|id),
                         data = df_long,
                         binwidth = 30,
                         family = "binomial",
                         pve = 0.95, npc = 4, periodicity = TRUE)
gcfpca_end_t = Sys.time()
gcfpca_time_diff = as.double(difftime(gcfpca_end_t, gcfpca_start_t, units="mins"))
gcfpca_time_diff
```

Here we will compare our model with GFAMM (Scheipl et al., 2015, 2016) using `pffr` functions in `refund` package. ${\rm logit}\{P\{Z_i(s)=1\}\} = \beta_0(s) + \beta_1(s) \text{Age}_i + \beta_2(s) \text{Gender}_i + b_i(s)$

```{r nhanes_pffr, warning=FALSE, message=FALSE}
I = length(unique(df_long$id))
K = length(unique(df_long$index))

df_pffr <- data.frame(Y = I(matrix(df_long$mims, I, K, byrow=TRUE)), 
                      id = factor(unique(df_long$id)), 
                      age = df_long$age[!duplicated(df_long$id)],
                      gender = df_long$gender[!duplicated(df_long$id)])

pffr_start_t = Sys.time()
pffr_model <- pffr(Y ~ age + gender + 
                     s(id, bs = "re", bs.yindex = list(k = 10, bs = "cr")), 
                 # use mgcv::bam with fastREML smoothing parameter 
                 # selection to estimate the model
                 algorithm="bam", method="fREML", discrete=TRUE,
                 # specify the bases used for estimating f_0(s), f_1(s) 
                 #via bs.int and bs.yindex, respectively
                 bs.yindex=list(bs = "cr", k=60),
                 bs.int=list(bs="cr", k=60),
                 data=df_pffr,
                 # specify outcome distribution and the functional domain (yind)
                 family="binomial", yind=1:1440)
pffr_end_t = Sys.time()
pffr_time_diff = as.double(difftime(pffr_end_t, pffr_start_t, units="mins"))

pffr_time_diff 

sind = 1:1440
df_pred <- data.frame("yindex.vec" = sind, 
                      id = as.numeric(levels(df_long$id))[1], 
                      age = 1, gender = 1)
betahat_pffr <- mgcv::predict.gam(pffr_model, newdata=df_pred, type='iterms', se.fit=TRUE)
CI_L_pffr <- betahat_pffr[[1]] - 1.96 * betahat_pffr[[2]]
CI_U_pffr <- betahat_pffr[[1]] + 1.96 * betahat_pffr[[2]]

eta_pffr <- matrix(pffr_model$linear.predictors, nrow = 50, ncol = 1440)
```

The plot below show the estimated fixed effect coefficients of the `GC-FPCA` and `GFAMM` models on the subsampled NHANES dataset

```{r plot_nhanes_fe, echo = TRUE, fig.show='hold'}
plot_df <- cbind.data.frame(sind = rep(seq(0, 1, length.out = K), 3),
                            betahat = c(data.matrix(nhanes_gcfpca$betaHat)),
                            betahat_pffr = c(data.matrix(betahat_pffr[[1]][, -4])),
                            X = c(rep("Intercept", K),
                                  rep("Age", K),
                                  rep("Gender", K)),
                            CI_L_pw = c(data.matrix(nhanes_gcfpca$CI_L_pw)),
                            CI_U_pw = c(data.matrix(nhanes_gcfpca$CI_U_pw)),
                            CI_L_joint = c(data.matrix(nhanes_gcfpca$CI_L_joint)),
                            CI_U_joint = c(data.matrix(nhanes_gcfpca$CI_U_joint)),
                            CI_L_pffr = c(CI_L_pffr[, -4]),
                            CI_U_pffr = c(CI_U_pffr[, -4])) %>%
  mutate(X = factor(X, levels = c("Intercept",
                                  "Age",
                                  "Gender")))

plot_df %>%
  ggplot(aes(x = sind, y = betahat)) +
  #geom_ribbon(aes(ymin = CI_L_joint, ymax = CI_U_joint, fill = "CI Joint"), alpha = 0.5) +
  geom_ribbon(aes(ymin = CI_L_pw, ymax = CI_U_pw, fill = "CI"), alpha = 0.3) +
  geom_ribbon(aes(ymin = CI_L_pffr, ymax = CI_U_pffr, fill = "CI pffr"), alpha = 0.5) +
  geom_line(aes(color = "GCFPCA")) +
  geom_line(aes(x = sind, y = betahat_pffr, color = "GFAMM")) +
  scale_fill_manual(values = c("CI" = "blue", "CI Joint" = "lightblue", "CI pffr" = "pink"), name = "Confidence Interval") +
  scale_color_manual(values = c("GCFPCA" = "blue", "GFAMM" = "red"), name = "Confidence Interval") +
  # Adding a horizontal dotted line at y = 0
  geom_hline(yintercept = 0, linetype = "dotted") +
  # Setting x-axis labels to show time
  scale_x_continuous(breaks = seq(0, 1, by = 0.2)) +
  # Facet the plot by variable X, with 3 columns
  facet_wrap(~X, ncol = 3, scales = "free_y") +
  # Adding axis labels and title
  labs(x = "Functional Domain", y = "", fill = "Confidence Interval")
```

The plot below shows latent subject-specific means on the linear predictor scale, ${\rm logit}\{P\{Z_i(s)=1\}\} = \beta_0(s) + \beta_1(s) \text{Age}_i + \beta_2(s) \text{Gender}_i + b_i(s)$, for three subjects in the subsampled NHANES dataset. GFAMM estimates are black solid lines while GCFPCA estimates are pink dotted lines.


```{r plot_nhanes_eta}
df_long %>%
  mutate(eta_hat = as.vector(nhanes_gcfpca$etas)) %>%
  mutate(eta_pffr = as.vector(eta_pffr)) %>%
  filter(id %in% c(63529, 67427, 82257)) %>%
  ggplot(aes(index, eta_pffr)) +
  geom_line() +
  geom_line(aes(y = eta_hat), linetype = 2, color = "salmon") +
  facet_wrap(~id)
```

The estimates from `GC-FPCA` and `GFAMM` are similar to each other for fixed effects and linear predictors.

The plot below show the estimated principal components for the GC-FPCA models. 

```{r plot_nhanes_re, echo = TRUE, fig.show='hold'}
phi_df = tibble(s = rep(seq(0, 1, length.out  = 1440), 4),
                l = rep(paste0("eigenfunction ", 1:4), each = 1440),
                logit_mod = c(data.matrix(nhanes_gcfpca$efunctions[, -1]))) %>%
  nest_by(l) 

phi_df %>%
  unnest(data) %>%
  ungroup() %>% 
  mutate(logit_mod = logit_mod * sqrt(1440)) %>%
  pivot_longer(logit_mod, names_to = "model", values_to = "value") %>%
  ggplot(aes(s, value, group = model, color = model, linetype = model)) +
  geom_line(color = "black") +
  facet_wrap(~l)
```

